An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-06-14 main.tex
Dagens text

Additionally, \Catra{} is able to efficiently generate a quantifier-free
Presburger formula form of such an instance of constraints and automata,
something beyond the reach of general model checkers like~\Nuxmv{}.

@@

where all variables $\TransitionVar_i, \StateVar_i, \FinalStateVar_i$ are existentially quantified and the variables
$\LetterVar_\Letter$ make up the actual image. Feeding this definition into a Presburger
solver and performing quantifier elimination on $\TransitionVar_i, \StateVar_i, \FinalStateVar_i$ with $\LetterVar_\Letter$ as free variables would produce the
Parikh image in its Presburger form.

In this paper we use this example (sometimes appropriately modified) as the
baseline approach. Notably, the approach presented in this paper can at its core
be described as lazily enforcing the connectedness constraint of this method
(the final two clauses) while also interleaving the (similarly lazy) computation
of products of automata.

\subsubsection{An Example}
We apply Equation~\ref{eq:generate-parikh} to the automaton of Figure~\ref{fig:len-branch}. We then get the following quantified Presburger formula. As before, we omit the implicit existential quantifiers for brevity, along with the constranints requiring all variables to be nonnegative. We also omit the $\land$-clauses between each line. In other words, in this example, only the variables $\LetterVar_a, \LetterVar_b, \LetterVar_c, \LetterVar_d, \LetterVar_e$, and $\LetterVar_f$ are free, and corresponds to the counts of the respective subscript letter. Note that since they here uniquely identify a transition due to the one-to-one mapping between character labels and transitions in the example, the first clause can be trivially eliminated by substituting $\LetterVar_a$ for $\TransitionVar_{\FromLabelTo{0}{a}{1}}$, etc. This is not generally the case.

\begin{equation}
  \begin{aligned}
  \ParikhMap(\Automaton):=\:
  &\LetterVar_{a} = \TransitionVar_{\FromLabelTo{0}{a}{1}} \land \LetterVar_{b} = \TransitionVar_{\FromLabelTo{0}{b}{2}} \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}} \land \LetterVar_{d} = \TransitionVar_{\FromLabelTo{2}{d}{3}} \land \LetterVar_{e} = \TransitionVar_{\FromLabelTo{2}{e}{2}} \land \LetterVar_{f} = \TransitionVar_{\FromLabelTo{1}{f}{3}}\\
  &\FinalStateVar_{3} > 0 \implies \StateVar_{3} > 0\\
  & \TransitionVar_{\FromLabelTo{0}{b}{2}} + \TransitionVar_{\FromLabelTo{0}{a}{1}} - 0 = 1 \land  \TransitionVar_{\FromLabelTo{1}{f}{3}} - \TransitionVar_{\FromLabelTo{0}{a}{1}} = 0\\
  & \TransitionVar_{\FromLabelTo{2}{d}{3}} - \TransitionVar_{\FromLabelTo{0}{b}{2}} = 0 \land \FinalStateVar_{3} + 0 - \TransitionVar_{\FromLabelTo{2}{d}{3}} - \TransitionVar_{\FromLabelTo{1}{f}{3}} = 0\\
  &\TransitionVar_{\FromLabelTo{0}{a}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} > 0 \implies \StateVar_{2} > 0\\
  &\TransitionVar_{\FromLabelTo{1}{c}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{2}{e}{2}} > 0 \implies \StateVar_{2} > 0\\
  &\TransitionVar_{\FromLabelTo{1}{f}{3}} > 0 \implies \StateVar_{3} > 0 \land \TransitionVar_{\FromLabelTo{2}{d}{3}} > 0 \implies \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \left(\StateVar_{0} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{0}{a}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{0} = \StateVar_{2} + 1 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} \geq 1 \land \StateVar_{2} \geq 1\right)\\
  &\StateVar_{1} > 0 \implies \left(\StateVar_{1} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{1}{c}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{1} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{1}{f}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{2} > 0 \implies \left(\StateVar_{2} = \StateVar_{2} + 1 \land \TransitionVar_{\FromLabelTo{2}{e}{2}} \geq 1 \land \StateVar_{2} \geq 1\right) \lor \left(\StateVar_{2} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{2}{d}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{3} > 0 \implies \FinalStateVar_{3} = 1
  \end{aligned}
  \end{equation}

@@

We implement a calculus for Parikh automata as described in
Section~\ref{sec:parikh-automata}. The artefact submitted along with this paper
is a program that reads one or more products of one or more DFA, the
register incrementations performed along the edges of their transitions, and the
labels of the transitions as ranges of Unicode characters, along with a set of
constraints on the final values of their registers expressed as Presburger
arithmetic. We call this program \Catra{}.\footnote{If you really must read it
as an acronym, please read it as CAtegory Theory on Register Automata, or
alternatively if you find this whole paper to be too much of a theoretical
exercise as Can Anyone Think of A Real Application.}

\Catra{} is written in Scala, with the calculus described in this paper
implemented as a theory plug-in for the \Princess{} automated theorem
prover~\cite{princess}, which also performs the Presburger reasoning. For
comparison, we also provide an implementation of the baseline method
from~\cite{generate-parikh-image}, a direct translation that uses the~\Nuxmv{}
symbolic model checker~\cite{nuxmv} to solve our constraints, and \Fudge{the
approximation described in~\cite{approximate-parikh} on top of the standard
Verma back-end}. An example of an input file \Fudge{illustrating every feature
of the file format} can be seen in Listing~\ref{lst:input-example}.

@@

\subsection{Implementing the Baseline}

@@

As a way of gauging the differences in overhead between the native backends and \Nuxmv{}, we executed a small experiment where we ran the same trivial instance 30 times and observed the runtimes for all solvers. All backends saw an improvement from JIT compilation within the first three runs, but subsequent runs only improved the native backends. The final measured overhead for the fully warm JVMs was less than~\OverheadSeconds{}. From this we draw the conclusion that major differences in runtime between the native backends and \Nuxmv{} is more likely to stem from innate differences rather than unfair overhead resulting from our particular configuration.

@@

It should be noted that despite running in the standard, deterministic configuration, we observed some degree of nondeterminism in \Nuxmv{}. Instances that were solved or unsolved would sometimes be solved or timed out in subsequent runs. This observation was made even without randomised execution order, suggesting that the nondeterminism likely stemmed either from our encoding of the instances into \Nuxmv{}'s format, the execution environment. Similarly, \Calculus{} showed similar signs of nondeterminism, but only for satisfiable instances. To address this, we increasded the runtime from our initial experiments to give us a wider marigin. By contrast, the baseline implementation showed no signs of nondeterminism. It should be noted that for both backends exhibiting nondeterminism, the nondeterminism did not change the general trend of the experiments, even for experiments with few instances. Therefore, we believe that our large set of benchmarks will protect us from bias since any backend is unlikely to be neither consistently lucky nor consistently unlucky.

>>>>

Additionally, \Catra{} is able to efficiently generate a quantifier-free
Presburger formula form of such an instance of constraints and automata,
something beyond the reach of general model checkers like~\Nuxmv.

@@

where all variables $\TransitionVar_i, \StateVar_i, \FinalStateVar_i$ are
existentially quantified and the free variables $\LetterVar_\Letter$ make up the
actual image. Performing standard quantifier elimination on $\TransitionVar_i,
\StateVar_i, \FinalStateVar_i$ with $\LetterVar_\Letter$ would produce the
Parikh image in its Presburger form. Note that compared to the method described
in~\cite{generate-parikh-image}, ours is backwards. Intuitively the approaches
are equivalent since the Parikh image, by definition, disregards order. The
approach can be described as either following paths through the automaton
forwards from an (in our case, \emph{the}) initial state, or conversely as in
the case here, choose an accepting state and trace a path backwards to the
initial state.

\contents{
  \item mention the bug here!
}

In this paper we refer to this model as the baseline approach, though we also
apply some optimisations that allow us to detect unsatisfiability early,
described in Section~\ref{sec:implementing-baseline}. Notably, the calculus
introduced in this paper can at its core be described as lazily enforcing the
connectedness constraint of this method (the final two clauses) while also
interleaving the (similarly lazy) computation of products of automata.

\subsubsection{An Example}
We apply Equation~\ref{eq:generate-parikh} to the automaton of Figure~\ref{fig:len-branch}. We then get the quantified Presburger formula seen in Equation~\ref{eq:parikh-formula}. As before, we omit the implicit existential quantifiers for brevity, along with the constranints requiring all variables to be nonnegative. We also omit the $\land$-clauses between each line. In other words, in this example, only the variables $\LetterVar_a, \LetterVar_b, \LetterVar_c, \LetterVar_d, \LetterVar_e$, and $\LetterVar_f$ are free, and corresponds to the counts of the respective subscript letter. Note that since they here uniquely identify a transition due to the one-to-one mapping between character labels and transitions in the example, the first clause can be trivially eliminated by substituting $\LetterVar_a$ for $\TransitionVar_{\FromLabelTo{0}{a}{1}}$, etc. This is not generally the case.

\begin{equation}
  \begin{aligned}
  &\LetterVar_{a} = \TransitionVar_{\FromLabelTo{0}{a}{1}} \land \LetterVar_{b} = \TransitionVar_{\FromLabelTo{0}{b}{2}} \land \LetterVar_{c} = \TransitionVar_{\FromLabelTo{1}{c}{1}} \land \LetterVar_{d} = \TransitionVar_{\FromLabelTo{2}{d}{3}} \land \LetterVar_{e} = \TransitionVar_{\FromLabelTo{2}{e}{2}} \land \LetterVar_{f} = \TransitionVar_{\FromLabelTo{1}{f}{3}}\\
  &\FinalStateVar_{3} > 0 \implies \StateVar_{3} > 0\\
  & \TransitionVar_{\FromLabelTo{0}{b}{2}} + \TransitionVar_{\FromLabelTo{0}{a}{1}} = 1 \land  \TransitionVar_{\FromLabelTo{1}{f}{3}} - \TransitionVar_{\FromLabelTo{0}{a}{1}} = 0\\
  & \TransitionVar_{\FromLabelTo{2}{d}{3}} - \TransitionVar_{\FromLabelTo{0}{b}{2}} = 0 \land \FinalStateVar_{3} + 0 - \TransitionVar_{\FromLabelTo{2}{d}{3}} - \TransitionVar_{\FromLabelTo{1}{f}{3}} = 0\\
  &\TransitionVar_{\FromLabelTo{0}{a}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} > 0 \implies \StateVar_{2} > 0\\
  &\TransitionVar_{\FromLabelTo{1}{c}{1}} > 0 \implies \StateVar_{1} > 0 \land \TransitionVar_{\FromLabelTo{2}{e}{2}} > 0 \implies \StateVar_{2} > 0\\
  &\TransitionVar_{\FromLabelTo{1}{f}{3}} > 0 \implies \StateVar_{3} > 0 \land \TransitionVar_{\FromLabelTo{2}{d}{3}} > 0 \implies \StateVar_{3} > 0\\
  &\StateVar_{0} > 0 \implies \left(\StateVar_{0} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{0}{a}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{0} = \StateVar_{2} + 1 \land \TransitionVar_{\FromLabelTo{0}{b}{2}} \geq 1 \land \StateVar_{2} \geq 1\right)\\
  &\StateVar_{1} > 0 \implies \left(\StateVar_{1} = \StateVar_{1} + 1 \land \TransitionVar_{\FromLabelTo{1}{c}{1}} \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{1} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{1}{f}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{2} > 0 \implies \left(\StateVar_{2} = \StateVar_{2} + 1 \land \TransitionVar_{\FromLabelTo{2}{e}{2}} \geq 1 \land \StateVar_{2} \geq 1\right) \lor \left(\StateVar_{2} = \StateVar_{3} + 1 \land \TransitionVar_{\FromLabelTo{2}{d}{3}} \geq 1 \land \StateVar_{3} \geq 1\right)\\
  &\StateVar_{3} > 0 \implies \FinalStateVar_{3} = 1
  \end{aligned}
  \label{eq:parikh-formula}
  \end{equation}

  We see here that the formula encodes the choice of path at the start of the
  automaton ($\TransitionVar_{\FromLabelTo{0}{b}{2}} +
  \TransitionVar_{\FromLabelTo{0}{a}{1}} = 1$), and the dependency between that
  choice and the following outgoing transition, e.g. in
  $\TransitionVar_{\FromLabelTo{2}{d}{3}} -
  \TransitionVar_{\FromLabelTo{0}{b}{2}} = 0$. Here we only have one choice of
  incoming flow; from $\FinalStateVar_3$, corresponding to the only accepting
  state. Simplifying the equation using standard algebra gives us this smaller
  set of clauses:

  \begin{equation}
    \begin{aligned}
    &\FinalStateVar_{3} > 0 \implies \StateVar_{3} > 0\\
    & \LetterVar_{d} + \LetterVar_{a} = 1\\
    & \FinalStateVar_{3}  > \LetterVar_{d} \implies \StateVar_{1} > 0 \land \LetterVar_{d} > 0 \implies \StateVar_{2} > 0\\
    &\LetterVar_{c} > 0 \implies \StateVar_{1} > 0 \land \LetterVar_{e} > 0 \implies \StateVar_{2} > 0\\
    &\FinalStateVar_{3}  > \LetterVar_{d} \implies \StateVar_{3} > 0 \land \LetterVar_{d} > 0 \implies \StateVar_{3} > 0\\
    &\StateVar_{0} > 0 \implies \left(\StateVar_{0} = \StateVar_{1} + 1 \land (\FinalStateVar_{3}  - \LetterVar_{d}) \geq 1 \land \StateVar_{1} \geq 1\right) \lor \left(\StateVar_{0} = \StateVar_{2} + 1 \land \LetterVar_{d} \geq 1 \land \StateVar_{2} \geq 1\right)\\
    &\StateVar_{1} > 0 \implies \left(\StateVar_{1} = \StateVar_{3} + 1 \land (\FinalStateVar_{3}  - \LetterVar_{d}) \geq 1 \land \StateVar_{3} \geq 1\right)\\
    &\StateVar_{2} > 0 \implies \left(\StateVar_{2} = \StateVar_{3} + 1 \land \LetterVar_{d} \geq 1 \land \StateVar_{3} \geq 1\right)\\
    &\StateVar_{3} > 0 \implies \FinalStateVar_{3} = 1
    \end{aligned}
    \label{eq:parikh-formula:reduced}
    \end{equation}

Note that some of the or-clauses disappear since they contain false
inequalities of the type $x = x + 1$, arising from cycles. These clauses encode
the distance in a spanning tree from an accepting state, and the
unsatisfiability is by design, essentially ensuring that two states in a cycle
cannot vouch for each other's connectedness.

Assume we are really interested in occurrences with the letter e, i.e. we solve
for $\LetterVar_e > 0$. This implies that $\StateVar_{2} > 0$ (essentially, that
the state our transition of interest started in is reachable), in turn implying
that $\StateVar_{2} = \StateVar_{3} + 1 \land \LetterVar_{d} \geq 1 \land
\StateVar_{3} \geq 1$. Armed with this knowledge, we now know that $\StateVar_3
> 0$, which means that $\FinalStateVar_3 = 1$. Since we also know that
$\LetterVar_d \geq 1$, we know that $\LetterVar_a = 0$. This corresponds to
choosing the bottom path through the automaton. Applying this to the equation
and performing standard boolean logic propagation gives the following:

\begin{equation}
  \begin{aligned}
  &\LetterVar_{d} = 1 \land \LetterVar_{a} = 0 \\
  & \StateVar_{3} > -1 \\
  &\LetterVar_{c} = 0\\
  &\StateVar_{0} > 0 \implies \StateVar_{0} = \StateVar_{3} + 2\\
  \end{aligned}
  \label{eq:parikh-formula:choose-a}
  \end{equation}

  Since the final implication holds (we know $\StateVar_3 \geq 0$, and therefore
  $\StateVar_3 + 2 > 0$), we can disregard it. Thereby we have eliminated all
  quantified variables and are left with a formula containing only the free
  ones. Note that all constraints on $\LetterVar_e$ have disappeared after we
  asserted that it was strictly positive. This is no coincidence: since it is a
  loop, it can take any positive value.

  This example, in addition to showing how the encoding works, has hopefully
  also illustrated one of the key insights behind the design of \Calculus: the
  fact that even few additional constraints on the Parikh image can dramatically
  reduce the difficulty of computing it, as in this extreme case where only one
  additional choice in practice fixes the values of all other symbols in the image.

@@

We implement a calculus for Parikh automata as described in
Section~\ref{sec:parikh-automata}. The artefact submitted along with this paper
is a program that reads one or more products of one or more DFA, the register
incrementations performed along the edges of their transitions, and the labels
of the transitions as ranges of Unicode characters, along with a set of
constraints on the final values of their registers expressed as Presburger
arithmetic. We call this program \Catra{}.\footnote{If you really must read it
as an acronym, please read it as CAtegory Theory on Register Automata, or if you
object to the somewhat nonstandard use of register automata and category theory,
as Check Assignments of The Registers Afterwards, or alternatively if you find it
all to be too much of a theoretical exercise as Can Anyone Think of A Real
Application.}

\Catra{} is written in Scala, with the calculus described in this paper
implemented as a theory plug-in for the \Princess{} automated theorem
prover~\cite{princess}, which also performs the Presburger reasoning. For
comparison, we also provide an implementation of the baseline method
from~\cite{generate-parikh-image}, a direct translation that uses the~\Nuxmv{}
symbolic model checker~\cite{nuxmv} to solve our constraints, and \Fudge{the
approximation described in~\cite{approximate-parikh} on top of the standard
baseline back-end}. An example of an input file \Fudge{illustrating every feature
of the file format} can be seen in Listing~\ref{lst:input-example}.

@@

\subsection{Implementing the Baseline}\label{sec:implementing-baseline}

@@

As a way of gauging the differences in overhead between the native backends and \Nuxmv, we executed a small experiment where we ran the same trivial instance 30 times and observed the runtimes for all solvers. All backends saw an improvement from JIT compilation within the first three runs, but subsequent runs only improved the native backends. The final measured overhead for the fully warm JVMs was less than~\OverheadSeconds. From this we draw the conclusion that major differences in runtime between the native backends and \Nuxmv{} is more likely to stem from innate differences rather than unfair overhead resulting from our particular configuration.

@@

It should be noted that despite running in the standard, deterministic configuration, we observed some degree of nondeterminism in \Nuxmv. Instances that were solved or unsolved would sometimes be solved or timed out in subsequent runs. This observation was made even without randomised execution order, suggesting that the nondeterminism likely stemmed either from our encoding of the instances into \Nuxmv's format, or the execution environment. Similarly, \Calculus{} showed similar signs of nondeterminism, but only for satisfiable instances. To address this, we increased the runtime from our initial experiments to give us a wider marigin. By contrast, the baseline implementation showed no signs of nondeterminism. It should be noted that for both backends exhibiting nondeterminism, the nondeterminism did not change the general trend of the experiments, even for experiments with few instances. Therefore, we believe that our large set of benchmarks will protect us from bias since any backend is unlikely to be neither consistently lucky nor consistently unlucky.
